Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-02 treestats_paper.tex
comments on algorithm

\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\indexin}[0]{\ensuremath{\mathcal{I}}}
\newcommand{\indexout}[0]{\ensuremath{\mathcal{O}}}

@@

\clearpage
\begin{taocpalg}{B}{General branch statistics}
{Given a list of $n$ sample nodes $S$, corresponding weights $w$ and
summary function $f: \R \rightarrow \R$
compute the span-normalised statistic $\sigma$
over a tree sequence with sequence length $L$
defined by the node and edge tables $\Nt$ and $\Et$. We assume
that the index vectors $\indexin$ and $\indexout$ have been precomputed.}

@@

\algstep{B2.}{Terminate.}{If $j \geq |\Et|$ or $t_l \geq L$
    set $\sigma \leftarrow \sigma / L$ and terminate.
}

\algstep{B3.}{Edge removal loop.}{If $k \geq |\Et|$ or $t_l \neq
    \Et[\indexout_k]\mbox{.right}$ go to \algref{B6}}.

\algstep{B4.}{Remove edge.}{Set
    $u \leftarrow \Et[\indexout_k]\mbox{.child}$,
    $v \leftarrow \Et[\indexout_k]\mbox{.parent}$ and $k \leftarrow k + 1$.
    Then set $s \leftarrow s - \beta_u F_u$, $\pi_u \leftarrow -1$
    and $\beta_u \leftarrow 0$.
}

@@

\algstep{B6.}{Edge insertion loop.}{If $j \geq |\Et|$ or $t_l \neq
    \Et[\indexin_j]\mbox{.left}$ go to \algref{B9}}.

\algstep{B7.}{Insert edge.}{Set
    $u \leftarrow \Et[\indexin_j]\mbox{.child}$,
    $v \leftarrow \Et[\indexin_j]\mbox{.parent}$ and $j \leftarrow j + 1$.
    Then set $\pi_u \leftarrow v$,
    $\beta_u \leftarrow \Nt[v]\mbox{.time} - \Nt[u]\mbox{.time}$ and
    $s \leftarrow s + \beta_u F_u$.
}

@@

\algstep{B9.}{Tree loop tail.}{Set $t_r \leftarrow L$.
    If $j < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexin_j]\mbox{.left})$.
    Then, if $k < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexout_k]\mbox{.right})$
    Finally, set $\sigma \leftarrow (t_r - t_l)s$,  $t_l \leftarrow t_r$
    and go to \algref{B2}.
}

>>>>

\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\prop}[1]{.\mbox{\texttt{#1}}} % property of a thing
\newcommand{\indexin}[0]{\ensuremath{\mathcal{I}}}
\newcommand{\indexout}[0]{\ensuremath{\mathcal{O}}}

@@

\plr{Question: Figures 1, 2 and 7 show the same sort of thing 
    (fig 1 is a site stat, aiming to give the general overview;
    figs 2 and 7 are branch stats). Should they be more similar?
    Do we need all of these?}

\plr{right now things like $\Et\mbox{.right}$ read like a funnily terminated sentence.
     I've changed these with a macro, feel free to adjust.}

\clearpage
\begin{taocpalg}{B}{General branch statistics}
{Given a list of $n$ sample nodes $S$, corresponding weights $w$ and
summary function $f: \R \rightarrow \R$
compute the span-normalised statistic $\sigma$
over a tree sequence with sequence length $L$
defined by the node and edge tables $\Nt$ and $\Et$. We assume
that the index vectors $\indexin$ and $\indexout$ have been precomputed.
\plr{I assume you plan to say in words what $\beta$, $x$, $F$, etc are}
}

@@

\algstep{B2.}{Terminate.}{If $j \geq |\Et|$ or $t_l \geq L$
    set $\sigma \leftarrow \sigma / L$ and terminate.
    \plr{perhaps say "return $\sigma/L$" to make it more clear what the output is?}
}

\algstep{B3.}{Edge removal loop.}{If $k \geq |\Et|$ or $t_l \neq
    \Et[\indexout_k]\prop{right}$ go to \algref{B6}.}

\algstep{B4.}{Remove edge.}{Set
    $u \leftarrow \Et[\indexout_k]\prop{child}$,
    $v \leftarrow \Et[\indexout_k]\prop{parent}$ and $k \leftarrow k + 1$.
    Then set $s \leftarrow s - \beta_u F_u$, $\pi_u \leftarrow -1$
    and $\beta_u \leftarrow 0$.
}

@@

\algstep{B6.}{Edge insertion loop.}{If $j \geq |\Et|$ or $t_l \neq
    \Et[\indexin_j]\prop{left}$ go to \algref{B9}.}

\algstep{B7.}{Insert edge.}{Set
    $u \leftarrow \Et[\indexin_j]\prop{child}$,
    $v \leftarrow \Et[\indexin_j]\prop{parent}$ and $j \leftarrow j + 1$.
    Then set $\pi_u \leftarrow v$,
    $\beta_u \leftarrow \Nt[v]\prop{time} - \Nt[u]\prop{time}$ and
    $s \leftarrow s + \beta_u F_u$.
}

@@

\algstep{B9.}{Tree loop tail.}{Set $t_r \leftarrow L$.
    If $j < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexin_j]\prop{left})$.
    Then, if $k < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexout_k]\prop{right})$
    Finally, set $\sigma \leftarrow (t_r - t_l)s$,  $t_l \leftarrow t_r$
    and go to \algref{B2}.
    \plr{I think you mean to add to $\sigma$, not replace it?}
}
